### 2019-11-13 paper/conceptual_framework.tex
First go at summary of spacing framework

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Whitworth2002,Molitoris2019}.

The mother may also be affected by the length of the birth intervals, although the
results on mother's anthropometric status are mixed \citep{Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.

@@

There are two effects of this theoretical results.
First, we should see an general increase in spacing across education groups to the extent that
development lead to a steeper income profile.
[This does require that economic development is associated with steeper income profiles;
not sure if anybody looked at this.
The lower fertility even for low education women does suggest that income is higher,
but does not, by itself, say anything about the age profile in income.]
Second, the increase in spacing should be stronger with increasing education levels.
In other words, we should see a general increase in spacing, and this increase should be
larger the higher the education level.

@@

@@

>>>>

One aspect of child ``quality'' where is is a strong effect of spacing is child health
and mortality, with most of the research finding that shorter spacing leads to worse
health and higher mortality for children, especially for very short intervals of 24 months 
or less \citep{Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
There is evidence in countries as diverse as Bangladesh, Brazil, El Salvadore, India,
and Pakistan of worse health outcomes and a higher mortality risk the shorter the birth
interval
\citep{Cleland1984,Curtis1993,Whitworth2002,Bhargava2003,Rutstein2005,Bhalotra2008,
Davanzo2008,Maitra2008,Makepeace2008,Gribble2009,Jayachandran2011,Saha2013,Jayachandran2017a,
Ghosh2018}.
This is different from the effect of the underlying mortality risk on birth spacing.
The expected effect of mortality risk---ignoring the effect of shorter spacing on mortality 
risk---is to shorten birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this type of effect 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize that having children close together increase the children's
mortality risk, they would prefer to space births further apart to lower mortality risk
(although possibly not very long since the mortality risk involves a trade-off between the
cost of births and the benefit of having a child).
Furthermore, the effects of short spacing on child mortality are strongly moderated by
maternal education, and even very short spacing has relatively low mortality for women
with secondary or tertiary education \citep{Whitworth2002,Molitoris2019}.

The mother may also be affected by the length of the birth intervals, although the
results on mother's anthropometric status are mixed and there appear to be no
effect on maternal mortality \citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
For India, there is, however, evidence that women with first-born girls are more likely to 
have anemia, possibly because short birth spacing can resulting in maternal depletion 
\citep{Milazzo2018}.

@@

There are two implications of this theoretical results.
First, we should see an general increase in spacing across education groups to the extent that
development lead to a steeper income profile.
[This does require that economic development is associated with steeper income profiles;
not sure if anybody looked at this.
The lower fertility even for low education women does suggest that income is higher,
but does not, by itself, say anything about the age profile in income.]
Second, the increase in spacing should be stronger with increasing education levels.
In other words, we should see a general increase in spacing, and this increase should be
larger the higher the education level.

@@

The question is how birth spacing is likely to change over time and with increasing
education in India.
As other countries like the US have developed, they have seen a progressively longer 
spacing between births as the proportion of births with very short intervals---less than 
24 months---have declined substantially \citep{Hotz1997}.
Most of the these countries do, however, not have the same degree of son preference that
we see in India.

What are the important factors that determine spacing?

- household income
- education
- labor force participation
- sex selection
- health (improvements, campaigns)

The main drivers of birth spacing of interest for India are female education and
development together with son preference and the increased availability of sex selection.

@@

\citep{Jensen2012}
\citep{Behrman2017}
\citep{Chatterjee2018}

On the U-shaped relationship between development and women's labor force participation \citep{Goldin1994}
